---
title: "Coding topic: tables"
output: html_document
---

```{r}
library(tidyverse)
```

The file "genus_info.csv" contains phenotype information for 200 bacterial
genera. The values were extracted from genus descriptions in the original
literature or in Bergey's Manual. Credit for the data set goes to Celeste
Gaughain, who completed the work as an undergraduate summer intern in 2017.

For the code exercise, I removed a few columns from the original data and
added a new column for the number of species in each genus. The number of
species is taken from the Living Tree Project release 128.

In the first code block, we read a CSV file into R using the function 
`read_csv()`. We use the left arrow `<-` to assign the data table to a new
variable named `genus_info`.

```{r}
genus_info <- read_csv("genus_info.csv")
```

Data tables occupy a central role in data analysis for science and statistics.
Data tables go by a few different names, like data frames, "tibbles", or simply
tables.

Data tables have rows and columns. Each row is supposed to correspond to one
"observation." Sometimes it's a little unclear what constitutes an observation,
but that's the idea you should keep in mind.

Each column represents something that you arranged or measured for the
observation. The values in each column are all of the same type. For example,
the `gram_stain` column contains text or character values. The `num_species`
column contains numbers.

We will cover five things that you can do with data tables. I think these five
activities will enable you to accomplish about 80% of the things you will ever
need to do with a data table.

## Filter rows with `filter()`

The first thing we'll need to do is to find observations that meet some
criteria. The `filter()` function does this. Let's find the rows where the
family is Enterobacteriaceae.

```{r}
genus_info %>%
  filter(family %in% "Enterobacteriaceae")
```

The command begins with the name of the table, `genus_info`. Then, we use a
pipe, `%>%`, to pass the table to the `filter()` function.

Inside the parentheses of the filter function, we write a condition that must
be met for each row to be retained. Notice that the column name, `family`, is
not in quotes, but the value, `"Enterobacteriaceae"`, is in quotes.

When filtering based on character columns, you'll often use the functions
`%in%` to specify allowed values or `str_detect()` to look for a pattern in
the values. When filtering based on numeric columns, you'll often use
mathematical expressions like `>`, `>=` (greater than or equal), `<`, or `<=`
(less than or equal). I recommend always using `%in%` when testing for
equality, for example, `filter(x %in% 3)`. It seems weird, but it will keep
you out of trouble. Finally, the exclamation point means "not," for example
`filter(!(family %in% "Enterobacteriaceae"))` would select everything *but*
the rows above.

## Select columns with `select()`

We can use the `select()` function to choose which columns to keep. This is
how we select for the columns `genus` and `gram_stain`.

```{r}
genus_info %>%
  select(genus, gram_stain)
```

## Create new columns with `mutate()`

We can create new columns in the table with `mutate()`.

To create new character columns, you'll often use the functions `str_replace()`
to replace patterns or `str_glue()` to "glue" together values from existing
columns.

```{r eval=FALSE}
genus_info %>%
  mutate(description = str_glue("{genus}, a {gram_stain} bacteria"))
```

To create new numeric columns, you'll often use matematical functions like `+`,
`-`, `*`, `/`, `log()`, or `sqrt()`.

## Compute aggregate values with `group_by()` and `summarize()`

We compute new columns with aggregate values in two steps. First, we group rows
in the table with `group_by()`. Then, we use `summarize()` to carry out the
computation within each group.

For example, here is how we compute the total number of species for each family
in the table:

```{r}
genus_info %>%
  group_by(family) %>%
  summarize(num_species_for_family = sum(num_species))
```

To summarize numeric columns, you'll often use functions like `sum()`,
`mean()`, `median()`, or `sd()`. The special function `n()` gives the number
of rows in each group, if you simply want to tally the rows.

## Count unique values with `count()`

Tallying the number of rows in each group is such a common operation that we
have a dedicated function for it: `count()`. Here is how we tabulate the number
of rows with each value of `gram_stain`:

```{r}
genus_info %>%
  count(gram_stain)
```

