---
output: html_document
---

```{r}
library(tidyverse)
```

```{r}
gg <- read_csv("greengenes_13_8_taxonomy.csv")
```

## Missing values

This table has a lot of missing values, which are coded as `NA`. Use the
function `is.na()` to filter based on missing values.

```{r}
gg %>%
  filter(!is.na(genus)) %>%
  filter(num_ref_seqs > 2000)
```

## Streptococcaceae

Let's see what else is in the Streptococcaceae family.

```{r}
gg %>%
  filter(family %in% "Streptococcaceae") %>%
  select(family, genus, num_ref_seqs)
```

## Phyla

Here are the total number of reference sequences by phylum.

```{r}
gg %>%
  group_by(phylum) %>%
  summarize(num_ref_seqs_for_phylum = sum(num_ref_seqs))
```

Wow, some reference sequences don't even have a phylum assignment. Let's look at the sequences that are missing a class assignment to get some more detail.

```{r}
gg %>%
  filter(is.na(class)) %>%
  select(kingdom, phylum, num_ref_seqs)
```

## Ruminococcus

We talked about Ruminococcus in the lecture. Let's see how many reference
sequences are filed under the *Ruminococcaceae* vs. the *Lachnospiraceae*
family. we have to use the `str_detect()` function to search for
"Ruminococcus", because the genus name might have brackets around it.

```{r}
gg %>%
  filter(str_detect(genus, "Ruminococcus")) %>%
  select(family, genus, num_ref_seqs)
```

