---
output: html_document
---

```{r}
library(tidyverse)
```

Here, we'll import the full set of species from the Living Tree Project,
revision 128.

```{r}
ltp <- read_csv("ltp_release_128_species.csv")
```

How many total species are in the table?

How many columns does the table have?

Filter the table to find only species in "Bacteroides."

```{r}
ltp %>%
  filter(____)
```

Select only the accession and species columns, in that order.

```{r}
ltp %>%
  select(____)
```

Count the number of species in each family.

```{r}
ltp %>%
  count(____)
```

Find the length of each 16S gene sequence from the start and end positions.

```{r}
ltp %>%
  mutate(seq_length = ____)
```

Make a label for each species that includes the family name, for example,
"Enterobacteriaceae - Escherichia coli".

```{r}
ltp %>%
  mutate(species_label = ____)
```

Find the median sequence length in each family.

```{r}
ltp %>%
  group_by(____) %>%
  summarize(median_seq_length = ____)
```

Filter the table to find the subspecies, that is, the species names that
contain "subsp".

```{r}
ltp %>%
  filter(____)
```
